A comparison on R tools for nonlinear least squares modeling

### 2023-03-04 RefactoringNLS/ComparingNLStools.Rmd
more added

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
task which is seemingly straightforward but turns out to be quite nasty. Because
this problem has a very succinct statement, it often provides the "short 
reproducible example" much beseeched of contributors to R mailing lists.

@@

Using a formula specification was one of the principal advances made
when `nls()` became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
In a one-parameter problem, think of $J$ as a very small number so that
we need to watch for a zero-divide or 0/0 situation in computing $\delta$.

@@

@@

`nlsr` tries to find solutions from very poor initial parameter values. While
this is not always possible, `nslr` has generally been able to succeed in finding
solutions, even when all parameters are started at 1. Let us compare its results
with those of `nls()` and `nlsLM()`.

@@

### Output object

The output object of `nlxb()` is smaller than that of `nls()` and is focused on 
the solution and its properties
rather than the **modeling** task. That is, package `nlsr` emphasizes the solution 
of the nonlinear least squares problem rather than the estimation of a nonlinear 
model that fits or explains the data. `nls()` and `nlsLM` return an object of 
class `nls`, which allows for
a number of specialized modeling and diagnostic extensions. For compatibility, 
the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
`nls` object. Unless particular modeling features are needed, the use of 
`wrapnlsr()` is unnecessary and wasteful of resources.

Internally, `nlxb()` uses symbolic and automatic differentiation tools to create 
the residual and Jacobian functions needed for `nlfb()` then uses it to solve 
the relevant nonlinear least squares minimization problem.

As with other nonlinear least squares packages, we provide the Jacobian code as 
the "gradient" attribute of the Jacobian function. This lets us embed the code 
for the Jacobian as this attribute of the **residual** function so that the call 
to `nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

Generally `nls()` and `nlsLM()` use a fairly simple forward difference 
approximation of derivatives for the Jacobian, though a central approximation 
can be specified in control parameters. Package `nlsr` provides four approximation 
options, with a further control `ndstep` for the size of the
step used in the approximation.

@@

## Feature: Convergence and termination tests (FIXED)

A previous issue with `nls()` that prevented it from providing parameter estimates for zero-residual
(i.e., perfect fit) data was corrected thanks to suggestions by one of us.

In the manual page for `nls()` in R 4.0.0 there is the warning:

>  The `nls` function uses a relative-offset convergence criterion
  that compares the numerical imprecision at the current parameter
  estimates to the residual sum-of-squares.  This performs well on data of
  the form

>  $$  y = f(x, \theta) + eps $$

>  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form

>  $$  y = f(x, \theta)  $$

>  because the criterion amounts to comparing two components of the round-off 
  error.

>  If you wish to test `nls` on artificial data please add a noise component, 
  as shown in the example below.

This amounted to admitting R cannot solve well-posed problems unless data is polluted
with errors.

This issue can be easily resolved. The "termination test" for the **program** rather than for 
"convergence" of the underlying **algorithm** is the Relative Offset Convergence Criterion 
(see @BatesWatts81). This projects the proposed step in the parameter vector on the gradient 
and estimates how much the sum of squares loss function should decrease. This estimate is divided 
by the current size of the loss function to avoid scale issues.
When we have "converged", the estimated
decrease is very small. However, with small residuals,
the sum of squares loss function is (almost) zero and we get the possibility of a 
zero-divide failure.

Adding a small quantity to the loss function before dividing avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
with a default value of zero. Setting it to a small number -- 1.0 is 
a reasonable choice --
allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
We are pleased to note that this improvement is now in the R distributed code since version 4.1.0.

<!-- ### Example of a small-residual problem -->

<!-- ``` -->
<!-- t <- -10:10 -->
<!-- y <- 100/(1 + .1 * exp(-0.51 * t)) -->
<!-- lform <- y ~ a/(1 + b * exp(-c * t)) -->
<!-- ldata <- data.frame(t = t, y = y) -->
<!-- plot(t, y) -->
<!-- lstartbad <- c(a = 1, b = 1, c = 1) -->
<!-- lstart2 <- c(a = 100, b = 10, c = 1) -->
<!-- nlsr::nlxb(lform, data = ldata, start = lstart2) -->
<!-- nls(lform, data = ldata, start = lstart2, trace = TRUE) -->
<!-- # Fix with scaleOffset -->
<!-- nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0)) -->
<!-- sessionInfo() -->
<!-- ``` -->

<!-- Edited output of as small example follows: -->

<!-- ``` -->
<!-- > t <- -10:10 -->
<!-- > y <- 100/(1 + .1 * exp(-0.51 * t)) -->
<!-- > lform <- y ~ a/(1 + b * exp(-c * t)) -->
<!-- > ldata <- data.frame(t=t, y=y) -->
<!-- > plot(t,y) -->
<!-- > lstartbad <- c(a = 1, b = 1, c = 1) -->
<!-- > lstart2 <- c(a = 100, b = 10, c = 1) -->
<!-- > nlsr::nlxb(lform, data = ldata, start = lstart2) -->
<!-- nlsr object: x  -->
<!-- residual sumsquares =  1.007e-19  on  21 observations -->
<!--     after  13    Jacobian and  19 function evaluations -->
<!--   name            coeff          SE       tstat      pval      gradient    JSingval    -->
<!-- a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6   -->
<!-- b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3   -->
<!-- c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791   -->
<!-- > nls(lform, data = ldata, start = lstart2, trace = TRUE) -->
<!-- 40346.    (1.08e+00): par = (100 10 1) -->
<!-- 11622.    (2.93e+00): par = (101.47 0.49449 0.71685) -->
<!-- 5638.0    (1.08e+01): par = (102.23 0.38062 0.52792) -->
<!-- 642.08    (1.04e+01): par = (102.16 0.22422 0.41935) -->
<!-- 97.712    (1.79e+01): par = (100.7 0.14774 0.45239) -->
<!-- 22.250    (1.78e+02): par = (99.803 0.093868 0.50492) -->
<!-- 0.025789  (1.33e+03): par = (100.01 0.10017 0.50916) -->
<!-- 6.0571e-08 (7.96e+05): par = (100 0.1 0.51) -->
<!-- 4.7017e-19 (1.86e+04): par = (100 0.1 0.51) -->
<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
<!-- ... (approx 40 lines omitted) -->
<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
<!-- Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) :  -->
<!--   number of iterations exceeded maximum of 50 -->
<!-- > # Fix with scaleOffset -->
<!-- > nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0)) -->
<!-- 40346.      (1.08e+00): par = (100 10 1) -->
<!-- 11622.      (2.91e+00): par = (101.47 0.49449 0.71685) -->
<!-- 5638.0      (9.23e+00): par = (102.23 0.38062 0.52792) -->
<!-- 642.08      (5.17e+00): par = (102.16 0.22422 0.41935) -->
<!-- 97.712      (2.31e+00): par = (100.7 0.14774 0.45239) -->
<!-- 22.250      (1.11e+00): par = (99.803 0.093868 0.50492) -->
<!-- 0.025789    (3.79e-02): par = (100.01 0.10017 0.50916) -->
<!-- 6.0571e-08  (5.80e-05): par = (100 0.1 0.51) -->
<!-- 4.7017e-19  (1.62e-10): par = (100 0.1 0.51) -->
<!-- Nonlinear regression model -->
<!--   model: y ~ a/(1 + b * exp(-c * t)) -->
<!--    data: ldata -->
<!--      a      b      c  -->
<!-- 100.00   0.10   0.51  -->
<!--  residual sum-of-squares: 4.7e-19 -->
<!-- Number of iterations to convergence: 8  -->
<!-- Achieved convergence tolerance: 1.62e-10 -->
<!-- ``` -->

### More general termination tests

The convergence criterion above, the principal termination control of `nls()`, leaves out 
some possibilities that could
be useful for some problems. The package `nlsr` (@nlsr2019manual) already offers both 
the safeguarded relative offset test (**roffset**) as well as a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
Unfortunately, the meaning of "iteration" varies among programs and requires 
careful examination of the
code. We prefer to use the number of times the residuals or the jacobian have
been computed and put upper limits on these. Our codes exit (terminate) when these limits
are reached. Generally we prefer larger limits than the default `maxiter = 50` of `nls()`,
but that may reflect the more difficult problems we have encountered because 
users consult us when standard tools have given unsatisfactory results.

>>>>

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
task which is seemingly straightforward but turns out to be quite nasty. Because
this problem has a very succinct statement, it often provides the "short 
reproducible example" much beseeched of contributors to R mailing lists. 
Nevertheless, this is not an artificially created test problem, but one that
came from a field researcher.

@@

Using a formula specification was one of the principal advances made
when `nls()` became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
In a one-parameter problem with ill-conditioning, think of $J$ as a 
very small number so that we need to watch for a zero-divide or 0/0 
situation in computing $\delta$.
The following show `nls()` does poorly on this example problem.

@@

Let us consider three alternatives to `nls()`: `nlsr::nlxb()`, `minpack.lm::nlsLM()`
and `gslnls::gsl_nls()`. We use default methods and controls in these examples.

### Solution attempts with nlsr

```{r ex02nlsr}
library(nlsr)
unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
print(unlx1) 
pshort(unlx1) # A short form output
snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
# pshort(snlx1)
print(snlx1)
tnlx1<-try(nlxb(formula=frmt, start=stt1, data=weeddf))
# pshort(tnlx1)
print(tnlx1)
```

### Solution attempts with minpack.lm

```{r ex02minpack}
library(minpack.lm)
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
summary(unlm1) # Use summary() to get display
unlm1 # Note the difference. Use this form to get sum of squares
# pnlm0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
# pnlm0(snlm1)
summary(snlm1)
## Following structure needed to deal with M1mac issue
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
if (inherits(tnlm1, "try-error")) {
   cat("Failure to compute solution -- likely singular Jacobian\n")
} else {  
   pnlm0(tnlm1) # short form to give sum of squares
   summary(tnlm1)
}   
```

### Solution attempts with gslnls

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
summary(ugslnls1) # Use summary() to get display
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
summary(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
summary(tgslnls1) # Use summary() to get display
```

### Comparison notes - formula setup

`nlsr::nlxb()` is set up to use `print()` to output standard errors and 
singular values of the Jacobian (for diagnostic purposes). By contrast, 
`minpack.lm::nlsLM()` and `nls()`
use `summary()`, which does NOT display the sum of squares, while `print()` 
gives the sum of squares, but not the standard error of the residuals.

The singular values displayed by `print.nlsr()` (the internal name for the 
adaptation of the generic `print()`) are displayed in a column to the right 
of the coefficient  and standard error display, but are NOT specific to the 
parameters. Their position is purely for efficient use of page space. The 
most common use of the singular values is to gauge how "nearly singular" the 
Jacobian is at the solution, and the ratio of the largest to smallest of the 
singular values is a simple but effective measure. In the above example, we note 
that the scaled logistic has the smallest ratio.

The results from `nlsLM` and `gsl_nls` for the transformed model have a very 
large sum of squares, which may suggest that these program have failed. 
Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
offer the singular values, we need to extract the Jacobian and compute its
singular values. The following script shows how to do this, using what is called
the `gradient` element in the returned solution for these solvers.

```{r ex04singval, echo=TRUE, eval=TRUE}  
# for nlsLM
if (inherits(tnlm1, "try-error"))  {
   print("Cannot compute solution -- likely singular Jacobian")
 } else {  
   JtnlsLM <- tnlm1$m$gradient() # actually the Jacobian
   svd(JtnlsLM)$d # Singular values
}   
# for gsl_nls
if (inherits(tgslnls1, "try-error")) {
   cat("Cannot compute solution -- likely singular Jacobian")
} else {  
   JtnlsLM <- tgslnls1$m$gradient()
   svd(JtnlsLM)$d # Singular values
}   
```

We see that there are differences in detail, but the more important result is that
two out of three singular values are essentially 0. Our Jacobian is singular, and no 
method of the Gauss-Newton type should be able to continue. Indeed, from this set of
parameters, `nlxb` also stops, even though the code in `nlsr` tries to find solutions
from very poor initial parameter values.

```{r ex05, echo=TRUE}  
stspecial<- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
badstart<-nlxb(formula=frmt, start=stspecial, data=weeddf)
print(badstart)
```

## Functional specification of problems

?? what do we want here??

## Design goals, termination tests and output objects

The output object of `nlxb()` is smaller than the class `nls` object returned
by `nls()`, `nlsLM()` and `gsl_nls()`. Package `nlsr` emphasizes the solution 
of the nonlinear least squares problem rather than the estimation of a nonlinear 
model that fits or explains the data. `nls()` and `nlsLM` return an object of 
class `nls`, which allows for a number of specialized modeling and diagnostic extensions. For compatibility, 
the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
`nls` object. Unless particular modeling features are needed, the use of 
`wrapnlsr()` is unnecessary and wasteful of resources.

The design goals of the different tools are also evident in the so-called
"convergence tests" for the iterative solvers. In the manual page for `nls()` 
in R 4.0.0 there was the warning:

and suggested adding small perturbations to the data. This amounted to 
admitting `nls()` cannot solve well-posed problems unless data is polluted
with errors. We dispute that zero-residual problems are always artificial,
since problems in function approximation and nonlinear equations can be 
approached with nonlinear least squares. Moreover, the matter can
can be easily resolved. The "termination test" for the **program** rather than for 
"convergence" of the underlying **algorithm** is the Relative Offset Convergence
Criterion (see @BatesWatts81). This projects the proposed step in the parameter
vector on the gradient and estimates how much the sum of squares loss function 
should decrease. This estimate is divided 
by the current size of the loss function to avoid scale issues.
When we have "converged", the estimated
decrease is very small. However, with small residuals,
the sum of squares loss function is (almost) zero and we get the possibility of a 
zero-divide failure.

Adding a small quantity to the loss function before dividing avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` with a default value of zero. Setting it to a small 
number -- 1.0 is a reasonable choice -- allows small-residual problems 
(i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**. The default value 
gives the legacy behaviour.
We are pleased to note that this improvement is now in the R distributed code 
since version 4.1.0, and it has been a part of `nlsr` since it was introduced.

More general termination tests can be used. `nlsr` a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

Termination after what may be considered excessive computation is also important.
`nls()` stops after `maxiter` "iterations".  
Unfortunately, the meaning of "iteration" varies among programs and requires 
careful examination of the
code. We prefer to use the number of times the residuals or the jacobian have
been computed and put upper limits on these. Our codes exit (terminate) 
when these limits are reached. Generally we prefer larger limits than the 
default `maxiter = 50` of `nls()`,
but that may reflect the more difficult problems we have encountered because 
users consult us when standard tools have given unsatisfactory results.

## Jacobian calculation

Gauss-Newton/Marquardt methods all need a Jacobian matrix at each iteration. 
By default, `nlsr::nlxb()` will try to evaluate this using analytic 
expressions using symbolic and automatic differentiation tools. 
When using formula specification of the model, `nls()`, `minpack.lm::nlsLM()`
and `gslnls::gsl_nls()` use a finite difference approximation to compute the
Jacobian, though `gsl_nls()` does have an option to attempt symbolic expressions.

When The Jacobian code as 
the "gradient" attribute of the output of the Jacobian function. This 
lets us embed the code 
for the Jacobian as this attribute of the **residual** function so that the call 
to `nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

Generally `nls()`, `nlsLM()` and `gsl_nls()` use a fairly 
simple forward difference approximation of derivatives for the Jacobian
from the base-R `stats` package called `numericDeriv()`. A central 
approximation can be specified in control parameters. 
Package `nlsr` provides, via appropriate calling syntax, four approximation 
options, with a further control `ndstep` for the size of the
step used in the approximation.

Analytic Jacobian code can be provided to `nls()`, `nlsLM()` and `gsl_nls()`,
and most `selfStart` models that automatically provide starting parameters
also include such code. There is documentation in R
of `selfStart`, but the construction of them is non-trivial. Their use with
`nlsr` is described in the "Introduction to nlsr" vignette. However, since `nlsr`
generally can use very crude starting values, we have rarely needed them.

@@

@@

@@
